### 2019-11-13 paper/paper.tex
Adding figures for results section of paper

The barrels are large cortical columns, identifiable from birth in layer 4 of
rodent primary somatosensory cortex as dense clusters of thalamocortical
axons, which are enclosed by borders a few neurons thick from postnatal day 3
\citep{erzurumlu_development_2012}. In the plane tangential to the cortical surface the
barrels constitute a map of the arrangement of the whiskers on the face, such
that the cells of adjacent barrels respond most strongly and most quickly to
deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. The barrels
reflect subcortical whisker maps comprising cellular aggregates known as
barrelettes in the brainstem and barreloids in the thalamus
\citep{killackey_pattern_1980}.

Each whisker and its corresponding barrelette, barreloid, and barrel can be
identified by its position on a two-dimensional grid
\citep{zucker_coding_1969,killackey_pattern_1980,van_der_loos_barreloids_1976}. The barrels are normally up to
twice as long in the axis aligned to the grid ‘rows’ compared to the ‘arcs’,
and the rows are visible from day 2 \citep{rebsam_refinement_2002} whereas the arcs become
distinct by day 3 \citep{erzurumlu_development_2012,rebsam_refinement_2002}. In metabotropic glutamate
receptor 5 (mGlu5) knockouts the barrel rows develop but the arcs do not
(\citealp{hannan_plc-1_2001} see also \citealp{fox_barrel_2008}), thus distinct processes
likely guide development of the barrels with respect to the orthogonal grid
axes.

@@

The boundaries of the barrels form polygonal shapes that together
create a Voronoi tesselation across the cortical sheet, wherein all
points enclosed by a given boundary are closer to its center than to
the center of any other boundary \citep{senft_mouse_1991}. This
patterning suggests that the barrel topology may be preserved in the
initial projection of the thalamocortical axons into the cortex, and
that a barrel forms by the lateral branching of the axons from their
initial centres of mass, which ceases upon contact with the axons
branching from adjacent centres. According to this model the barrel
pattern could be entirely determined by the distribution of the
initial centre points, which would itself be determined by guidance
cues. However, the assumption of pre-arranged centres is difficult to
resolve with the observation that thalamocortical axons arrive in the
cortical plate as an undifferentiated bundle prior to barreloid
formation (see \citealp{erzurumlu_development_2012} for a review).

@@

\cite{karbowski_model_2004} developed what remains, to the best of our knowledge, the
only dynamical model of cortical arealization capable of generating
predictions about the influence of signalling molecules on the size, shape,
and topological organization of emergent cortical fields. This model describes
competition between thalamocortical projections for territory along a
one-dimensional string of cortical cells, e.g., along an anterior-posterior
axis. The equations of the model define how the density of connections, $c$,
and the density of axon branches, $a$, interact at time $t$ at each cortical
location $x$, for $N$ thalamocortical projections indexed by $i$. The Voronoi
tesselation used to describe barrel patterning is a two-dimensional structure,
hence we first extended the original model for simulation on a two-dimensional
cortical sheet, and describe it here as such (i.e., substituting $x$ with the
two-dimensional position vector $\mb{x}$ and substituting terms of the form
$\frac{d}{dx}$ with the two-dimensional spatial gradient operator $\nabla$).

@@

\cmnt{If we need to present the results of the 2-D system with $\chi=0$, then
  we'll do that first, showing a 2-D version of the simulation obtained by
  running \code{james0 configs/karb.json}}

We first verified that the system given by Eqns. \ref{eq:dc} \&
\ref{eq:da} described a two dimensional extension of the axon
branching model in \citet{karbowski_model_2004}. We showed that in the presence of
three guidance gradients varying only in one dimension (along the long axis of
an elliptical domain), five thalamocortical axon types, each interacting
uniquely with the three guidance gradients, would segregate into five separate
regions. We investigated a number of configurations of the guidance gradients,
varying their shape, $\rho(\mb{x})$, and number, $M$, and their effect on a
varying number of thalamocortical projections, $N$. In general, we were not
able to produce structure in the resulting patterns of the connections,
$c(\mb{x},t\rightarrow\infty)$, which was not present in the guidance
patterns.

Complex structure can indeed be observed in protein expression for some genes,
such as the molecules Bhlhb5 \citep{joshi_bhlhb5_2008} and Lmo4
\citep{cederquist_lmo4_2013,greig_molecular_2013}. These expression patterns can therefore
provide a theoretical mechanism for the guidance of axons into distinct
regions. The question of how such structures are generated becomes one of of
how gene transcriptional interactions and mRNA diffusion and movement achieves
this (\cmnt{Fixme: find a ref on structure in gene transcription}). However,
complex and stable structure in gene expression data is commonly reported;
monotonically varying expression patterns, such as those for Pax6, Emx2, Sp8
and Coup-tf1 which display graded expression across the entire isocortex
(\cmnt{REFS}) are more typical. Furthermore, the complex structure observed in
the whisker barrel field has not, to our knowledge, been linked to a complex
set of gene expression patterns matching the scale of the barrels. To
investigate how the whisker barrel structure might form in the presence of
only monotonically varying morphogen gradients, we turned our attention to the
competition for connections built into the model.

\subsection*{Competition is required to form somatotopically ordered barrels}

@@

This introduces a new parameter, $\epsilon$, which governs the
strength of the branching competition. The factor $1/(N-1)$ ensures that the
branching competition strength scales with the number of TC axons in the
system.

@@

%
\begin{equation} \label{eq:branch_comp_pow_logistic}
\chi(\mb{x}, t) = - \frac{\epsilon a_i}{N-1} \sum_{i' \ne i}^{N}
\frac{1}{1 + e^{-m (a_{i'})^l}} - \frac{1}{2}
\end{equation}
%
$\epsilon$ represents the maximum value of the competition term for TC
type $i$. The parameter $m$ controls the steepness of the function.

\cmnt{Start from the thalamic barreloids, to generate the interaction
  parameters}

% Include figure ../boundaries/barreloids_vanderloos_graph.png
\begin{figure*}
\begin{center}
\includegraphics[width=0.7\textwidth]{../boundaries/barreloids_vanderloos_graph.png}
\end{center}
%\vspace{-0em}
\caption{Barreloid centres taken from \cite{van_der_loos_barreloids_1976}, Fig.~2
  are shown as black circles. It is assumed that there are at least
  two linearly increasing, orthogonal signalling gradients in the
  thalamus (side graphs - addme!), such that the location of each
  barreloid within the graded thalamic signal field can differentiate
  the efferent axons, with the result that axons from each barreloid
  interact to varying degrees with cortical signalling
  gradients. Coloured arrows show, for each barreloid, the magnitude
  of the interaction that its thalamocortical projection will have
  with the four signalling gradients in the cortex. The mappings from
  barreloid location to interaction strength are linear.}
\label{fig:barreloids}
\end{figure*}

>>>>

The barrels are large cortical columns, identifiable from birth in layer 4 of
rodent primary somatosensory cortex as dense clusters of thalamocortical
axons, which are enclosed by borders a few neurons thick from postnatal day 3
\citep{erzurumlu_development_2012}. In the plane tangential to the cortical surface the
barrels constitute a map of the arrangement of the whiskers on the
face (see \citealp{yamakado_subdivision_1979}), such
that the cells of adjacent barrels respond most strongly and most quickly to
deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. The barrels
reflect subcortical whisker maps comprising cellular aggregates known as
barrelettes in the brainstem and barreloids in the thalamus
\citep{killackey_pattern_1980}.

Each whisker and its corresponding barrelette, barreloid, and barrel can be
identified by its position on a two-dimensional grid
\citep{zucker_coding_1969,killackey_pattern_1980,van_der_loos_barreloids_1976,haidarliu_size_2001}. The barrels are normally up to
twice as long in the axis aligned to the grid ‘rows’ compared to the ‘arcs’,
and the rows are visible from day 2 \citep{rebsam_refinement_2002} whereas the arcs become
distinct by day 3 \citep{erzurumlu_development_2012,rebsam_refinement_2002}. In metabotropic glutamate
receptor 5 (mGlu5) knockouts the barrel rows develop but the arcs do not
(\citealp{hannan_plc-1_2001} see also \citealp{fox_barrel_2008}), thus distinct processes
likely guide development of the barrels with respect to the orthogonal grid
axes.

@@

The boundaries of the barrels form polygonal shapes that together
create a Voronoi tesselation across the cortical sheet, wherein all
points enclosed by a given boundary are closer to its center than to
the center of any other boundary \citep{senft_mouse_1991}. This
patterning suggests that the barrel topology may be preserved in the
initial projection of the thalamocortical axons into the cortex, and
that a barrel forms by the lateral branching of the axons from their
initial centres of mass, which ceases upon contact with the axons
branching from adjacent centres. According to this model the barrel
pattern could be entirely determined by the distribution of the
initial centre points, which would itself be determined by guidance
cues. However, the assumption of pre-arranged centres is difficult to
resolve with the observation that thalamocortical axons arrive in the
cortical plate as an undifferentiated bundle prior to barreloid
formation (\citealp{agmon_organized_1993}; see \citealp{erzurumlu_development_2012} for a review).

@@

\cite{karbowski_model_2004} developed what remains, to the best of our knowledge, the
only dynamical model of cortical arealization capable of generating
predictions about the influence of signalling molecules on the size, shape,
and topological organization of emergent cortical fields. This model describes
competition between thalamocortical projections for territory along a
one-dimensional string of cortical cells, e.g., along an anterior-posterior
axis. The equations of the model define how the density of connections, $c$,
and the density of axon branches, $a$, interact at time $t$ at each cortical
location $x$, for $N$ thalamocortical projections indexed by $i$. The Voronoi
tesselation used to describe barrel patterning is a two-dimensional structure,
hence we first extended the original model for simulation on a two-dimensional
cortical sheet, and describe it here as such (i.e., substituting $x$ with the
two-dimensional position vector $\mb{x}$ and substituting terms of the form
$\frac{d}{dx}$ with the spatial gradient operator $\nabla$).

@@

%\cmnt{If we need to present the results of the 2-D system with $\chi=0$, then
%  we'll do that first, showing a 2-D version of the simulation obtained by
%  running \code{james0 configs/karb.json}}

We first verified that the system given by Eqns. \ref{eq:dc} \& \ref{eq:da}
described a two dimensional extension of the axon branching model in
\citet{karbowski_model_2004}. We showed that in the presence of three guidance
gradients varying only in one dimension (along the long axis of an elliptical
domain), five thalamocortical axon types, each interacting uniquely with the
three guidance gradients, would segregate into five separate regions. We
investigated a number of configurations of the guidance gradients, varying
their shape, $\rho(\mb{x})$, and number, $M$, and their effect on a varying
number of thalamocortical projections, $N$. In general, we were not able to
producen a complexity of structure in the resulting patterns of the
connections, $c(\mb{x},t\rightarrow\infty)$, which was not present in the
guidance patterns.

Complex structure can indeed be observed in protein expression for some genes,
such as the molecules Bhlhb5 \citep{joshi_bhlhb5_2008} and Lmo4
\citep{cederquist_lmo4_2013,greig_molecular_2013}. These expression patterns
can therefore provide a theoretical mechanism for the guidance of axons into
distinct regions.
%The question of how such structures are generated becomes
%one of of how gene transcriptional interactions and mRNA diffusion and
%movement achieves this (\cmnt{Fixme: find a ref on structure in gene
%  transcription}).
However, complex and stable structure in gene expression data is not commonly
reported; monotonically varying expression patterns, such as those for Pax6,
Emx2, Sp8 and Coup-tf1, which display graded expression across the entire
isocortex (\cmnt{REFS}) are more typical. Furthermore, the complex structure
observed in the whisker barrel field has not, to our knowledge, been linked to
a set of gene expression patterns matching the scale of the barrels. To
investigate how the whisker barrel structure might form in the presence of
only monotonically varying morphogen gradients, we turned our attention to the
competition for connections built into the model.

\subsection*{Branching competition is required to form somatotopically ordered barrels}

@@

This introduces a new parameter, $\epsilon$, which governs the strength of the
branching competition. The factor $1/(N-1)$ ensures that the branching
competition strength scales with the number of TC axon types in the system.

We first investigate the effect of the competition by examining a system of 25
thalamocortical axon types simulated on a square domain. The axons interact
with two orgthogonal signalling gradients, one of which increases from the top
edge to the bottom edge, the other of which increases from the right edge to
the left edge. The values of $\gamma$ are chosen as a grid.

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{./Fig_25N2M_square/Fig_25N2M_square.png}
\end{center}
%\vspace{-0em}
\caption{\textbf{(a)--(e)} Connection densities of 25 thalamocortical axons
  simulated on a square domain. \textbf{(a)} Connection density at simulation
  time t=0. The left image shows a map of the identity of the thalamocortical
  axon type with the highest absolute connection density. Hexagonal elements
  coloured the darkest blue have the highest connection density for TC type
  $i=1$; those coloured the darkest red have maximal connection density in TC
  type $i=25$. The right hand image of each pair shows the hexes for which the
  connection density passes 0.5 for each TC type, giving a contour map. At
  t=0, connection densities are randomly distributed. \textbf{(b)} t=0.14. The
  grid-like organisation of the TC connections is beginning to emerge.
  \textbf{(c)} t=0.2. The 5x5 grid is now clear in the left map. Branching
  competition and differential interaction with the two guidance gradients is
  causing the branching, and therefore the connection densities, to localise in
  25 different locations. \textbf{(d)} t=0.64. Connection densities for each
  TC type have increased sufficiently for the contours to form. In (d), the
  mapping between colour and TC type $i$ is indicated. \textbf{(e)}
  t=1.41. Contours are well formed. In (e), the $\gamma$ values for each TC
  type, $i$, are shown. $\gamma_{i,1}$ gives the interaction with molecule `1'
  [red map in (f)]; $\gamma_{i,2}$ is the interaction with molecule
  `2'. \textbf{(f)} The two guidance gradients (`1' and `2') for this
  simulation.}
\label{fig:square}
\end{figure*}

\subsection*{The model can transfer realistic barrel patterns}

In which we show that a realistic barrel pattern in the thalamus can be
recreated in barrel cortex.

\subsubsection*{Stronger branching competition}

\cmnt{The realistic barrel pattern experiment is enhanced by stronger
  branching competition. This section may be required here or elsewhere.}

@@

%
\begin{equation} \label{eq:branch_comp_pow_logistic}
\chi(\mb{x}, t) = - \frac{\epsilon a_i}{N-1} \sum_{i' \ne i}^{N}
\frac{1}{1 + e^{-m (a_{i'})^l}} - \frac{1}{2}
\end{equation}
%
$\epsilon$ represents the maximum value of the competition term for TC type
$i$. The parameters $l$ and $m$ control the steepness and shape of the
function.

%%% End of stronger branching competition %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}
\begin{center}
\includegraphics[width=0.7\textwidth]{../boundaries/barreloids_haidarliu_graph.png}
\end{center}
%\vspace{-0em}
\caption{Barreloid centres taken from \cite{haidarliu_size_2001} are shown as
  black circles. It is assumed that there are two linearly increasing,
  orthogonal signalling gradients (molecules A \& B) in the thalamus (side
  graphs), such that the location of each barreloid within the graded thalamic
  signal field can differentiate the efferent axons, with the result that
  axons from each barreloid interact to varying degrees with cortical
  signalling gradients. Coloured arrows show, for each barreloid, the
  magnitude (and sign) of the interaction that its thalamocortical projection
  will have with the signalling gradients in the cortex. The mappings from
  barreloid location to interaction strength are linear.}
\label{fig:barreloids}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{./realistic_barrels.png}
\end{center}
%\vspace{-0em}
\caption{Realistic barrel pattern}
\label{fig:barrels}
\end{figure*}
